---
title: "Getting & Cleaning Data Course Project"
author: "Aharon"
date: "7/5/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Codebook

The goal today is to transform the Human Activity Recognition Using Smartphones Data Set from the UCI Machine Learning Repository into a more readable, more easily understandable, summarized format. We can begin by noting the features and observations as they were displayed in the original dataset, before describing the transformations each variable underwent.

## Feature Transformations
### Original Features
The original features of the dataset are as follows:

* tBodyAcc-XYZ
* tGravityAcc-XYZ
* tBodyAccJerk-XYZ
* tBodyGyro-XYZ
* tBodyGyroJerk-XYZ
* tBodyAccMag
* tGravityAccMag
* tBodyAccJerkMag
* tBodyGyroMag
* tBodyGyroJerkMag
* fBodyAcc-XYZ
* fBodyAccJerk-XYZ
* fBodyGyro-XYZ
* fBodyAccMag
* fBodyAccJerkMag
* fBodyGyroMag
* fBodyGyroJerkMag

In each case, the 't' stands for time, the 'f' stands for frequency domain signals, 'Body' refers to body movement, 'Gravity' refers to gravitational movement, 'Acc' refers to acceleration, 'Jerk' refers, as best I understand it, to jerking motions, and the 'Mag' refers to the magnitude of each motion. 

Further, each of the above features is given a wide variety of summary variables to help explain it. Everything from mins and maxes to autoregression coefficients are included. For the purpose of this assignment, I'll be pulling out the average and standard deviation for each feature, with the X, Y, and Z variables separated to make alternative calculations easier for whoever uses this dataset. 
      
### Transformed Features
Outside of simply subsetting the above variables so that only the averages and standard deviations were included, I changed the variable names in various ways to increase legibility. The new variable names are as follows:
      
* BodyAccel_Time-XYZ
* GravityAccel_Time-XYZ
* BodyAccel_TimeJerk-XYZ
* BodyGyro_Time-XYZ
* BodyGyro_TimeJerk-XYZ
* BodyAccel_TimeMag
* GravityAccel_TimeMag
* BodyAccel_TimeJerkMag
* BodyGyro_TimeMag
* BodyGyro_TimeJerkMag
* BodyAccel_Freq-XYZ
* BodyAcc_FreqJerk-XYZ
* BodyGyro_Jerk-XYZ
* BodyAccel_FreqMag
* BodyAccel_FreqJerkMag
* BodyGyro_FreqMag
* BodyGyro_FreqJerkMag

As already mentioned, each of these features is, in the final dataset, summarized via averages and standard deviations, written as "avg" and "std" within the datasets variable names.
The transformations shown above increase clarity in two ways: first, by making explicit that "t" is for "time" and "f" is for "frequency", and second by adding underscores to more easily differentiate between the types of measurement being described. 
 
      
## Transforming Selected Observations
### Prevoius observations, need for improvement
Over the course of the project, several of the variable names needed to be revised for clarity. I'm referring specifically to the activity names which, in the original dataset, are stored as numbers in the following way:
      
* 1 = WALKING
* 2 = WALKING_UPSTAIRS
* 3 = WALKING_DOWNSTAIRS
* 4 = SITTING
* 5 = STANDING
* 6 = LAYING

To better communicate these observations in the final, tidy dataset, a new column with the acitvity names written explicitly was created to replace the activity numbers of the original dataset. 

Further, numerous calculations were made for each patient (labeled in the "subject" field), which had to be averaged. This was accomplished using the "group_by" function, allowing, in the final dataset, for each observation to be of a particular subject, within a particular activity
      
## The Cleaning Process

This section will explain how I achieved the transformations described above, and created a clean, viable dataset for future use.
      
I began by pulling in the test and training datasets, along with the other files necessary for complete analysis within each. Example code for the test dataset is as follows:
      
```{r}
library(dplyr)
##Pull in test data
test_raw <- read.table(file = "./UCI HAR Dataset/test/X_test.txt",
                   header = FALSE)
##Pull in test data feature names
y_test <- read.table(file = "./UCI HAR Dataset/test/y_test.txt",
                   header = FALSE)
##Pull in test data subject identifier
test_subject <- read.table(file = "./UCI HAR Dataset/test/subject_test.txt",
                       header = FALSE)

###Pulling in training data
##Pull in training data feature names
y_train <- read.table(file = "./UCI HAR Dataset/train/y_train.txt",
                      header = FALSE)
##Pull in training data
train_raw <- read.table(file = "./UCI HAR Dataset/train/X_train.txt",
                        header = FALSE)
##Pull in train data subject identifier
train_subject <- read.table(file = "./UCI HAR Dataset/train/subject_train.txt",
                            header = FALSE)
```

I then pulled in the features dataset common to both training and testing datasets with the following code:
      
```{r}
features <- read.table(file = "./UCI HAR Dataset/features.txt",
                       header = FALSE)
```

I then made some minor name changes to the original data to make variable transformations easier later on, though this step seems exceptionally optional. 
      
```{r}
##Renaming test and train data
colnames(test_raw) <- features$V2
colnames(train_raw) <- features$V2

##Renaming test and train labels
y_test <- rename(y_test, activity = "V1")
y_train <- rename(y_train, activity = "V1")

##Renaming subject label
test_subject <- rename(test_subject, subject = "V1")
train_subject<- rename(train_subject, subject = "V1")
```
      
      
Finally, I combined each of the datasets with both the feature name and  subject files.
      
```{r}
##Combining test and labels and subjects
test_data <- cbind(test_subject, y_test, test_raw)

##Combining train and labels and subjects
train_data <- cbind(train_subject, y_train, train_raw)
```

I then combined the test and train datasets as follows:
      
```{r}
data <- rbind(test_data, train_data)
```

However, given that I needed only the average and standard deviation measurements for each of the necessary variables, I simply pulled out the required variables and placed them into a newdataset to simplify later transformations
      
```{r}
##Combing training and testing data
data <- rbind(test_data, train_data)
datatemp1 <- grep("mean\\()", names(data))
datatemp2 <- grep("std", names(data))
data2 <- data[,c(datatemp1, datatemp2)]
```

But can't forget the subject and activity columns! So I just bind those onto the newdataset without a problem:

```{r}
data2 <- cbind(data[,1:2], data2)
```

Next come the observation name changes for activities. I took the long way to do this but, using a character vector with the proper names would work just as well.
      
```{r}
data2$activityname[data2$activity == 1] <- "walking"
data2$activityname[data2$activity == 2] <- "walking_upstairs"
data2$activityname[data2$activity == 3] <- "walking_downstairs"
data2$activityname[data2$activity == 4] <- "sitting"
data2$activityname[data2$activity == 5] <- "standing"
data2$activityname[data2$activity == 6] <- "laying"

data2 <- select(data2, -activity)
```

I then moved this new variable to the front of the dataset, removed the old activity variable, and moved on to the final transformations. 

First, I need to use group_by on the subject column to more easily average the results of each test for each subject and each exercise.
      
```{r}
data2 <- data2[, c(1, 68, 2:67)]

data3 <- data2 %>%
      group_by(subject, activityname) %>%
      summarize_all("mean")
```

Then comes the long and annoying process of changing the feature names, which I did as follows:
      
```{r}
newnames <- gsub("tBodyAcc", "BodyAccel_Time", names(data3))
newnames <- gsub("tGravityAcc", "GravityAccel_Time", newnames)
newnames <- gsub("tBodyGyro", "BodyGyro_Time", newnames)
newnames <- gsub("fBodyAcc", "BodyAccel_Freq", newnames)
newnames <- gsub("fBodyGyro", "BodyGyro_Freq", newnames)
newnames <- gsub("fBodyBodyGyro", "BodyGyro_Freq", newnames)
newnames <- gsub("fBodyBodyAcc", "BodyAccel_Freq", newnames)
newnames <- gsub("-mean()", "_avg", newnames)
newnames <- gsub("\\()", "", newnames)

data4 <- data3 %>%
      rename_at(names(data3), ~ newnames)
```

And that's that! The final dataset contains separate columns for the averaged results of each subject and each test. Took a little while, but that's one tidy dataset. 
